---
title: "Fungi Network analysis"
author: "Francesco Delogu"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

# Intro

## Load libraries and setup

```{r load libraries, results='hide', message=F, warning=F}
options(stringsAsFactors=F)

library(tidyverse)
library(WGCNA)
library(igraph)
library(arules)
library(rlist)

wd <- getwd()
```

## Load data and preprocessing

```{r load data, results='hide', message=F, warning=F}
raw_data <- read.table(paste0(wd, "/data/raw_perseus.txt"), header=T, sep="\t", quote = "")

df.thermo <- raw_data %>%
  filter(Organism=="m.thermophila") %>%
  select(Uniprot, Bagasse1, Bagasse2, Bagasse3, Birch1, Birch2, Birch3, Spruce1, Spruce2, Spruce3, Glucose1, Glucose2, Glucose3) %>%
  column_to_rownames(., var="Uniprot")

df.chryso <- raw_data %>%
  filter(Organism=="p.chrysosporium") %>%
  select(Uniprot, Bagasse1, Bagasse2, Bagasse3, Birch1, Birch2, Birch3, Spruce1, Spruce2, Spruce3, Glucose1, Glucose2, Glucose3) %>%
  column_to_rownames(., var="Uniprot")

df.terreus <- raw_data %>%
  filter(Organism=="a.terreus") %>%
  select(Uniprot, Bagasse1, Bagasse2, Bagasse3, Birch1, Birch2, Birch3, Spruce1, Spruce2, Spruce3, Glucose1, Glucose2, Glucose3) %>%
  column_to_rownames(., var="Uniprot")

df.jecor <- raw_data %>%
  filter(Organism=="h.jecorina") %>%
  select(Uniprot, Bagasse1, Bagasse2, Bagasse3, Birch1, Birch2, Birch3, Spruce1, Spruce2, Spruce3, Glucose1, Glucose2, Glucose3) %>%
  column_to_rownames(., var="Uniprot")

df.crassa <- raw_data %>%
  filter(Organism=="n.crassa") %>%
  select(Uniprot, Bagasse1, Bagasse2, Bagasse3, Birch1, Birch2, Birch3, Spruce1, Spruce2, Spruce3, Glucose1, Glucose2, Glucose3) %>%
  column_to_rownames(., var="Uniprot")

Prot2type <- raw_data$Enzyme.type
names(Prot2type) <- raw_data$Uniprot
```

# Co-expression networks

## Network function

```{r make.net function}

make.net <- function(expr) {
  expr[is.na(expr)] <- (-10)
  adj <- adjacency(t(expr), type="unsigned")
  #sort(adj.thermo[upper.tri(adj.thermo)], decreasing=T)[(dim(adj.thermo)[1]**2-dim(adj.thermo)[1])/2/10]
  return(adj)
}
```

## Infer networks

```{r make networks}

adj.thermo <- make.net(df.thermo)
adj.chryso <- make.net(df.chryso)
adj.terreus <- make.net(df.terreus)
adj.jecor <- make.net(df.jecor)
adj.crassa <- make.net(df.crassa)

```

# Enzyme type networks

## Second network function

```{r make.enzyme.net function}

make.enzyme.net <- function(adj, enzyme.names){
  enzymes <- unlist(str_split(Prot2type[colnames(adj)][Prot2type[colnames(adj)]!=""], ";"))
  enzyme.net <- matrix(rep(0, length(enzyme.names)**2), nrow=length(enzyme.names))
  colnames(enzyme.net) <- enzyme.names
  rownames(enzyme.net) <- enzyme.names
  for(i in enzyme.names){
    selected_i <- unlist(lapply(str_split(Prot2type[colnames(adj)], ";"), function(x){i%in%x}))
    for(j in enzyme.names){
      selected_j <- unlist(lapply(str_split(Prot2type[colnames(adj)], ";"), function(x){j%in%x}))
      tot <- table(enzymes)[i]*table(enzymes)[j]
      if(!is.na(tot)){
        if(i==j){
          if(tot-table(enzymes)[i]==0){
            enzyme.net[i, j] <- NA
          } else {
            enzyme.net[i, j] <- (sum(adj[rownames(adj)[selected_i], colnames(adj)[selected_j]])-table(enzymes)[i])/(tot-table(enzymes)[i])
          }
        } else {
          enzyme.net[i, j] <- sum(adj[rownames(adj)[selected_i], colnames(adj)[selected_j]])/tot
        }
      } else {
        enzyme.net[i, j] <- NA
      }
    }
  }
  return(enzyme.net)
}

```

## Infer enzyme netorks

```{r make enzyme networks}

enzyme.names <- unique(unlist(str_split(Prot2type[Prot2type!=""], ";")))

enzymes.thermo <- make.enzyme.net(adj.thermo, enzyme.names)
enzymes.chryso <- make.enzyme.net(adj.chryso, enzyme.names)
enzymes.terreus <- make.enzyme.net(adj.terreus, enzyme.names)
enzymes.jecor <- make.enzyme.net(adj.jecor, enzyme.names)
enzymes.crassa <- make.enzyme.net(adj.crassa, enzyme.names)

```

## Plot enzyme netorks

```{r plot enzyme nets}

corrplot::corrplot(enzymes.thermo)
corrplot::corrplot(enzymes.chryso)
corrplot::corrplot(enzymes.terreus)
corrplot::corrplot(enzymes.jecor)
corrplot::corrplot(enzymes.crassa)

```

## export cytoscape individual networks

```{r cytoscape 5 nets}

cyt = exportNetworkToCytoscape(enzymes.thermo, edgeFile = paste0(wd, "/results/CytoscapeInput-edges-", "thermo", ".txt"), nodeFile = paste0(wd, "/results/CytoscapeInput-nodes-", "thermo", ".txt"), weighted=T, threshold=0, nodeNames=colnames(enzymes.thermo))
cyt = exportNetworkToCytoscape(enzymes.chryso, edgeFile = paste0(wd, "/results/CytoscapeInput-edges-", "chryso", ".txt"), nodeFile = paste0(wd, "/results/CytoscapeInput-nodes-", "chryso", ".txt"), weighted=T, threshold=0, nodeNames=colnames(enzymes.chryso))
cyt = exportNetworkToCytoscape(enzymes.terreus, edgeFile = paste0(wd, "/results/CytoscapeInput-edges-", "terreus", ".txt"), nodeFile = paste0(wd, "/results/CytoscapeInput-nodes-", "terreus", ".txt"), weighted=T, threshold=0, nodeNames=colnames(enzymes.terreus))
cyt = exportNetworkToCytoscape(enzymes.jecor, edgeFile = paste0(wd, "/results/CytoscapeInput-edges-", "jecor", ".txt"), nodeFile = paste0(wd, "/results/CytoscapeInput-nodes-", "jecor", ".txt"), weighted=T, threshold=0, nodeNames=colnames(enzymes.jecor))
cyt = exportNetworkToCytoscape(enzymes.crassa, edgeFile = paste0(wd, "/results/CytoscapeInput-edges-", "crassa", ".txt"), nodeFile = paste0(wd, "/results/CytoscapeInput-nodes-", "crassa", ".txt"), weighted=T, threshold=0, nodeNames=colnames(enzymes.crassa))

```

# General network (constrained)

## All enzymes network

```{r make enzyme total networks}

enzyme.tot <- matrix(rep(NA, length(enzyme.names)**2), nrow=length(enzyme.names))
colnames(enzyme.tot) <- enzyme.names
rownames(enzyme.tot) <- enzyme.names
for(i in enzyme.names){
  for(j in enzyme.names){
    result <- mean(c(enzymes.thermo[i,j], enzymes.chryso[i,j], enzymes.terreus[i,j],
                              enzymes.jecor[i,j], enzymes.crassa[i,j]), na.rm=T)
    enzyme.tot[i,j] <- ifelse(is.nan(result), NA, result)
  }
}
corrplot::corrplot(enzyme.tot)

```

# Plots I

## Igraph

```{r igraph plot}

enzyme.tot[is.na(enzyme.tot)] <- 0

g <- graph_from_adjacency_matrix(abs(enzyme.tot), mode="undirected", weighted=T)
E(g)$width <- E(g)$weight*10

l <- layout_with_fr(g)
plot(g, edge.arrow.size=.4, vertex.shape="none", layout=l)

```

## Cytoscape

```{r export network}

enzyme.tot[is.na(enzyme.tot)] <- 0

cyt = exportNetworkToCytoscape(enzyme.tot, edgeFile = paste0(wd, "/results/CytoscapeInput-edges-", "Fungi", ".txt"), nodeFile = paste0(wd, "/results/CytoscapeInput-nodes-", "Fungi", ".txt"), weighted=T, threshold=0, nodeNames=colnames(enzyme.tot))

```

## Export protein counts

```{r export protein counts}

summ_data <- raw_data %>%
  filter(Enzyme.type!="") %>%
  select(Enzyme.type, Organism) %>% 
  separate_rows(Enzyme.type, sep = ";", convert = TRUE) %>%
  group_by(Enzyme.type, Organism) %>%
  summarise(N=n()) %>%
  spread(Organism, N)
summ_data[is.na(summ_data)] <- 0
summ_data <- cbind(as.data.frame(summ_data), tot=rowSums(summ_data[,2:6]))

write.table(summ_data, file=paste0(wd, "/results/Enzyme_counts.txt"), row.names=F, quote=F, sep="\t")

```

# Significance of the edges

## Discretization

```{r discretize adjacency}

adj.discrete <- discretize(enzyme.tot, method="interval", labels=F, breaks=101, include.lowest=T)-1
adj.discrete <- matrix(adj.discrete, ncol=sqrt(length(adj.discrete)))
colnames(adj.discrete) <- colnames(enzyme.tot)
rownames(adj.discrete) <- rownames(enzyme.tot)
write.csv(adj.discrete, file=paste0(wd, "/results/enzyme_tot.csv") , quote=F)

```

## Significance functions

```{r significance.fun}

indirect.edge.test <- function(w, ku, kv, q){
  # q must be half of the sum of all the edges on the graph
  p <- (ku*kv)/(2*q**2)
  s <- binom.test(w, q, p, alternative="greater")$p.value
  return(s)
}

MLF <- function(adj){
  
  adj[is.na(adj)] <- 0
  degs <- colSums(adj) # you may want remove the diagonal
  q <- round(sum(degs)/2)
  
  new_adj <- matrix(NA, nrow=nrow(adj), ncol=ncol(adj))
  rownames(new_adj) <- rownames(adj)
  colnames(new_adj) <- colnames(adj)
  
  for(i in rownames(adj)){
    for(j in colnames(adj)){
      new_adj[i,j] <- indirect.edge.test(adj[i,j], degs[i], degs[j], q)
    }
  }
  return(new_adj)
}

```

## Compute significance 

```{r compute significance}

enzymes.Ps <- MLF(adj.discrete)

enzymes.Ps <- ifelse(enzymes.Ps<0.05, 1, 0)
enzymes.pruned <- enzyme.tot * enzymes.Ps

```

# Plots II

## Pruned graph plot

```{r pruned igraph plot}

g_pruned <- graph_from_adjacency_matrix(enzymes.pruned, mode="undirected", weighted=T)
E(g_pruned)$width <- E(g_pruned)$weight*10

l <- layout_with_fr(g_pruned)
plot(g_pruned, edge.arrow.size=.4, vertex.shape="none", layout=l)

```

## Cytoscape pruned

```{r export pruned network}

enzymes.pruned[is.na(enzymes.pruned)] <- 0

cyt = exportNetworkToCytoscape(enzymes.pruned, edgeFile = paste0(wd, "/results/CytoscapeInput-edges-", "Fungi_pruned", ".txt"), nodeFile = paste0(wd, "/results/CytoscapeInput-nodes-", "Fungi_pruned", ".txt"), weighted=T, threshold=0, nodeNames=colnames(enzymes.pruned))

```

## Degree distribution plot

```{r plot degrees}

degs <- sort(strength(g), decreasing=T)
ggplot(data=NULL, aes(y=degs, x=factor(names(degs), levels=names(degs)))) +
  geom_bar(stat="identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x="Enzyme type", y="Strength", title="Strength Distribution")

```

